---
title: "Google Data Analytics Capstone project"
author: "Marc Tagne"
date: '2022-07-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
setwd("C:/Users/marco/Desktop/coursera/Google data analytics course/capstone_project/2021_01_2022_05_divvy-tripdata")

```

## installing packages

## this package is already installed, hence it is commented out
```{r, include = False}
# install.packages("RSQLite")
```



## loading necessary packages
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(RSQLite)
library(writexl)
```

## loading 12 month worth of data on cyclistic data trips

```{r}

df1<-read.csv("202101-divvy-tripdata.csv")
df2<-read.csv("202102-divvy-tripdata.csv")
df3<-read.csv("202203-divvy-tripdata.csv")
df4<-read.csv("202204-divvy-tripdata.csv")
df5<-read.csv("202205-divvy-tripdata.csv")
df6<-read.csv("202106-divvy-tripdata.csv")
df7<-read.csv("202107-divvy-tripdata.csv")
df8<-read.csv("202108-divvy-tripdata.csv")
df9<-read.csv("202109-divvy-tripdata.csv")
df10<-read.csv("202110-divvy-tripdata.csv")
df11<-read.csv("202111-divvy-tripdata.csv")
df12<-read.csv("202112-divvy-tripdata.csv")

```

## checking the structure of the files loaded
```{r}
str(df1)
str(df2)
str(df3)
str(df4)
str(df5)
str(df6)
str(df7)
str(df8)
str(df9)
str(df10)
str(df11)
str(df12)
```

## by checking the structure it is clear all the loaded csv filles have 13 variables and differ in length. It is therefore possible to combine/merge them all together. For this,there are a number of ways we could use bind_rows() or rbind(). here i use bind_rows from the tidyverse package
```{r}
cyclic_data<-bind_rows(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)

# now i removed the loaded dataframes because of memory issues
rm(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)

# removing empty columns  and rows

cyclic_data<- janitor::remove_empty(cyclic_data,which = c("cols"))
cyclic_data<- janitor::remove_empty(cyclic_data, which = c("rows"))


```

## now checking the cyclic_data
```{r}
head(cyclic_data)

str(cyclic_data)

colnames(cyclic_data)
```

## from the data it can be observed that the columns start_at and end_at  are character types and formated as date time they should be seperated. 
```{r, results='hide'}
# preparing the data for analysis
cyclic_data2<-cyclic_data

# converting dates and time from the start our into POSIXct characters. For references about POSIxct see https://www.neonscience.org/resources/learning-hub/tutorials/dc-convert-date-time-posix-r

cyclic_data2$started_at<-as.POSIXct(cyclic_data2$started_at, format ="%Y-%m-%d %H:%M:%S",tz="GMT")

cyclic_data2$ended_at<-as.POSIXct(cyclic_data2$ended_at, format ="%Y-%m-%d %H:%M:%S", tz="GMT")

#checking for na values in the data set
apply(is.na(cyclic_data2),2,which)

# I can see that there are a few na values observed e.g.:

(cyclic_data2[4575733,])
(cyclic_data[4575733,])

# creating new columns for hours, days, year and day of the week columns.
##


df <- data.frame(DateTime = seq(as.POSIXct("2012-01-01"),
                                as.POSIXct("2012-02-01"), by=(30*60)))

df$hour<-hour(df$DateTime)
##


cyclic_data2$date<-as.Date(cyclic_data2$started_at)
cyclic_data2$month <-format(as.Date(cyclic_data2$date), "%b")

cyclic_data2$day <-format(as.Date(cyclic_data2$date), "%d")
cyclic_data2$year <-format(as.Date(cyclic_data2$date), "%Y")
cyclic_data2$day_of_week <-weekdays(cyclic_data2$date)

# this is not needed but i want to represent weekdays into numbers in an nother column with 1 representing monday and so on. t

cyclic_data2<- cyclic_data2 %>% mutate(day_of_week_2 = case_when(
  grepl("Monday", day_of_week) ~ 1,
  grepl("Tuesday", day_of_week) ~ 2,
  grepl("Wednesday", day_of_week) ~ 3,
  grepl("Thursday", day_of_week) ~ 4,
  grepl("Friday", day_of_week) ~ 5,
  grepl("Saturday", day_of_week) ~ 6,
  grepl("Sunday", day_of_week) ~ 7))


```

## basic data cleaning 

## should have been done before
```{r, warning=FALSE, results='hide'}

# basic cleaning cyclic data
head(cyclic_data2)

#removing cyclic_data due to memory issues
rm(cyclic_data)

# checking that the columns rideable_type and member_casual contain the appropriate information. member_casual should contain either member or casual and rideable_type contains either docked_bike, electric_bike or classic_bike

unique(cyclic_data2$rideable_type)

unique(cyclic_data2$member_casual)

# we are now sure we have the appropriate information in both columns

# checking for na values
apply(is.na(cyclic_data2), 2, which) # here we see that the column end_lat, end_lng have a lot of na values.

# we will drop the na values using the drop_na() function as this will become a problem later on if we don't remove them

cyclic_data3<-drop_na(cyclic_data2)

#removing cyclic_data2 due to memory issues
rm(cyclic_data2)

## adding the columns start_hour and end_hour with the help of lubridate package function "hour"

cyclic_data3$start_hour<-lubridate::hour(cyclic_data3$started_at)
cyclic_data3$end_hour<-lubridate::hour(cyclic_data3$ended_at)

# calculate trip duration
#cyclic_data3$ride_length_hours<-difftime(cyclic_data3$ended_at, cyclic_data3$started_at, units = "hours")

# should be ride length in hours in stead of Hours

#test_2$mins<-difftime(test_2$ended_at,test_2$started_at, units = "mins")


# calculating the time spent riding bikes (ride_length) i.e the difference in time between started_at and ended_at. 

cyclic_data3$ride_length_minutes<-difftime(cyclic_data3$ended_at,cyclic_data3$started_at, units = "mins") # ridelength in minutes

#checking for NA values
apply(is.na(cyclic_data3),2,which)


# droping minutes values that are less than or equal to zero (<=0). we will only work with values greater than zero in the ride_length_minutes column

cyclic_data4<-cyclic_data3 %>% filter(ride_length_minutes>0) %>% drop_na()

#we remove cyclic_data3 now
rm(cyclic_data3)


```

# before ploting the graphs a few things need to be adjusted 
```{r}
unique(cyclic_data4$day_of_week)
unique(cyclic_data4$month)
cyclic_data4$day_of_week <- factor(cyclic_data4$day_of_week, levels = c("Monday","Tuesday", "Wednesday", "Thursday", "Friday","Saturday","Sunday") )
cyclic_data4$month<-factor(cyclic_data4$month, levels = c("Jan","Feb", "Mar","Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))


cyclic_data4$ride_length_minutes<-as.numeric(cyclic_data4$ride_length_minutes)

# removing stations_names_with blank
cyclic_data4<-cyclic_data4 %>%filter(start_station_name != "")


# will save this for safety purposes
# write_csv(cyclic_data4,"./cyclic_data4.csv")
```

## Creating summary data frame and plots

## finding the percentage of casual riders and members in the dataset
```{r}
percen_member_casual<-cyclic_data4 %>% group_by(member_casual) %>% summarise(count=length(ride_id), percentage = (length(ride_id)/nrow(cyclic_data4)) * 100)

print(percen_member_casual)

# we can see that there are more members(56.3%) than casual riders(43.7%)

# representing with pie chart 
ggplot(percen_member_casual,aes(x = "", y = percentage, fill = member_casual))+geom_bar(stat ="identity", width = 1) + coord_polar("y", start = 0)+ guides(fill =guide_legend(title = "membership") )+theme_void() + labs(title = "Membership percentage")
```
## Finding the most rides per weekdays
```{r}
#most most rides per week
most_rides_week<-cyclic_data4 %>%
  group_by(day_of_week) %>%
  summarise(average_ride_length = mean(ride_length_minutes),
            number_of_rides = n())

ggplot(most_rides_week, aes(x = day_of_week, y=number_of_rides))+geom_col( color = "blue", fill = "blue")+theme_bw() + labs(title = "Number of rides per week", x = "\n Weekday", y = "Number of rides \n")

# graph shows that the most rides per week are on weekends(saturdays and sundays)
```

## Determining the number of rides by membership per week
```{r}
# determining the ride length by membership per week
ride_length_per_membership<-cyclic_data4 %>%
  group_by(member_casual,day_of_week) %>%
  summarise(average_ride_length = mean(ride_length_minutes),
            number_of_rides = n())

# representing with the number of rides per week days
ggplot(ride_length_per_membership,aes(x = day_of_week, y=number_of_rides, fill = member_casual))+geom_col(position = "dodge")+labs(title="Number of rides per weekdays per membership", x = "\n Weekday", y = "Number of rides\n")+theme_bw()+
  theme(axis.text.x = element_text(angle = 45),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))

# the graph clearly shows that we have more casual members on weekend

```

## representing the average ride length in minutes per week days
```{r}

ggplot(ride_length_per_membership,aes(x = day_of_week, y= average_ride_length, fill = member_casual))+geom_col(position = "dodge")+labs(title="Ride length in minutes per weekdays by membership", x = "\n Weekday", y = "ride length in minutes\n")+theme_bw()+
  theme(axis.text.x = element_text(angle = 45),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))

#the graph shows that the average ridelength per week for casual riders is higher than that of members 

```
## determing the number of rides per month
```{r}
#ride length per month
ride_length_per_month_only<-cyclic_data4 %>%
  group_by(month) %>%
  summarise(average_ride_length = mean(ride_length_minutes),
            tot_ride_length = sum(ride_length_minutes),
            number_of_rides = n())

# representing the number of rides per month
ggplot(ride_length_per_month_only,aes(x = month, y = number_of_rides ))+geom_col(position="dodge",color = "darkred", fill = "red")+ labs(title = "Number of rides per month", x ="\n Months", y = "Number of rides \n") +guides(fill =guide_legend(title = "Membership"))+theme_bw()

# the graph shows that the months of June to september have the highest number of rides


```
## ride length per membership per month
```{r}

ride_length_per_month<-cyclic_data4 %>%
  group_by(member_casual,month) %>%
  summarise(average_ride_length = mean(ride_length_minutes),
            tot_ride_length = sum(ride_length_minutes),
            number_of_rides = n())

# representing the number of rides per membership per month
ggplot(ride_length_per_month,aes(x = month, y = number_of_rides, fill = member_casual))+geom_col(position="dodge")+ labs(title = "Number of rides per month by membership", x ="\n Months", y = "Number of rides \n") +guides(fill =guide_legend(title = "Membership"))

# representing the average ride length per membership per month
ggplot(ride_length_per_month,aes(x = month, y = average_ride_length, fill = member_casual))+geom_col(position="dodge")+ labs(title = "Averge ride lengths in minutes per month by membership", x ="\n Months", y = "Average ride length in minutes \n") +guides(fill =guide_legend(title = "Membership"))+ theme_bw()

# representing the total ride length per month
ggplot(ride_length_per_month,aes(x = month, y = tot_ride_length, fill = member_casual))+geom_col(position="dodge")+ labs(title = "total ride lengths in minutes per month by membership", x ="\n Months", y = "Total ride length in minutes \n") +guides(fill =guide_legend(title = "Membership")) 

# The first graph shows something suprising, casual riders have the highest number of rides from of june to august
# The second graphs shows that casual riders have the highest average riding time per month throughout the year i.e have more riding time
# The third graph shows that the number of casual riders have the highest total riding time frpm march to august


```
# Finding the number of rides per starting hours each week per membership. This will help identify at what time riders are the most active
```{r}
# finding the number of rides by starting hours by week per membership
bike_rides_week_hours<-cyclic_data4 %>% group_by(day_of_week,start_hour,member_casual) %>%
  summarise(
    Minutes_total = sum(ride_length_minutes),
    Average = mean(ride_length_minutes),
    number_of_rides = n()
  ) %>% ungroup()

# representing the number of rides by hours
ggplot(bike_rides_week_hours, aes(x = start_hour, y =number_of_rides, fill = member_casual))+geom_col() + labs(title = "Number of rides by hours per week", x ="\n Hours", y="Number of rides\n" )+ guides(fill =guide_legend(title = "Membership"))

# representing the Number of rides by hours for day of the week by membership
ggplot(bike_rides_week_hours, aes(x = start_hour, y =number_of_rides, fill = member_casual))+geom_col() + labs(title = "Number of rides by hours per week", x ="\n Hours", y="Number of rides\n" )+facet_wrap(~day_of_week)+guides(fill =guide_legend(title = "Membership"))

# First graph shows that riders are most active between 3 and 8pm for both casual and active riders
# The second graph shows that clear differences in pattern between weekdays and weekends. interesting! on weekends there is more activity between 9 am and 8pm. WHy? this requires more investigation.

```

# Finding the number of rides by starting hours by month per memberships
```{r}
# number of rides by starting hours by month per membership
bike_rides_month_hours<-cyclic_data4 %>% group_by(month,start_hour,member_casual) %>%
  summarise(
    Minutes_total = sum(ride_length_minutes),
    Average = mean(ride_length_minutes),
    number_of_rides = n()
  ) %>% ungroup()

# representing the Number of rides by hours per month by membership
ggplot(bike_rides_month_hours, aes(x = start_hour, y =number_of_rides, fill = member_casual))+geom_col() + labs(title = "Number of rides by hours per month", x ="\n Hours", y="Number of rides\n" )+facet_wrap(~month)+guides(fill =guide_legend(title = "Membership"))

```

# Finding the most popular bike types per weekday for each membership type
```{r}
# Finding the most popular bike types per week for each membership
bike_type_week<-cyclic_data4 %>% group_by(day_of_week,rideable_type,member_casual) %>%
  summarise(
    Minutes_total = sum(ride_length_minutes),
    Mean = mean(ride_length_minutes),
    number_of_rides = n()
  ) %>% ungroup()

# representing the most popular bike types per week for each memberships
ggplot(bike_type_week,aes(x = day_of_week,y= number_of_rides, fill = member_casual)) + geom_col()+facet_wrap(~rideable_type)+labs(title ="Most popular types of bikes per membership per week", x = "\n Week days", y = "Number of rides\n")+guides(fill =guide_legend(title = "Membership"))+
  theme(axis.text.x = element_text(angle = 45),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))

# the data shows clearly that docked_bike types are the least favorite bike types while the most favorite one is classic bike follwed by electric bike. 

```

# Finding the most popular bike types per memberships
```{r}
# Finding the most popular bike types per memberships
bike_type<-cyclic_data4 %>% group_by(rideable_type,member_casual) %>%
  summarise(
    Minutes_total = sum(ride_length_minutes),
    average_ride_time = mean(ride_length_minutes),
    number_of_rides = n()
  ) %>% ungroup()

# representing the most popular bike types per memberships
ggplot(bike_type, aes(x=rideable_type, y =number_of_rides, fill=member_casual,))+geom_col(position = "dodge")+labs(title = "most popular bike types per memberships", x = "\n bike types", y = "Number of rides\n" )+guides(fill =guide_legend(title = "Membership"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))
```

#Finding out who rides more on days of the week and weekend for each membership type
```{r}
# finding out who rides more on days of the week and weekend by memberships
more<-cyclic_data4 %>% group_by(member_casual,day_of_week) %>% 
  summarise(
    average_ride_length = mean(ride_length_minutes),
    number_of_rides = n()
  ) %>% ungroup()

more<-more %>% mutate(type_of_week = ifelse(day_of_week == "Saturday"| day_of_week == "Sunday", "Weekend", "Midweek"))

ggplot(more, aes(x = member_casual, y = number_of_rides, fill = member_casual))+ geom_col()+ labs(title = "count of riders on weekdays and weekends", x = "\n membership", y ="number of rides \n")+ facet_wrap(~type_of_week)+guides(fill =guide_legend(title = "Membership"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))
```
#finding the percentage of bike types, ride length, number of rides, average ride length and total ride length
```{r}
# finding the mmost popular biketypes 
bike_type_month<-cyclic_data4 %>% group_by(rideable_type) %>%
  summarise(
    Minutes_total = sum(ride_length_minutes),
    average_ride_time = mean(ride_length_minutes),
    number_of_rides = n()) %>% ungroup %>%  
  mutate(total= sum(number_of_rides),
         percentage = (number_of_rides/total) *100)
print(bike_type_month)

#
```

# Finding  the most start stations
```{r}
# finding the top stations(i.e start and end stations)
length(unique(cyclic_data4$start_station_name))
length(unique(cyclic_data4$start_station_id))


# finding the most busiest stations per memberships
bike_stations<-cyclic_data4 %>% group_by(start_station_name) %>%
  summarise(
    Minutes_total = sum(ride_length_minutes),
    average_ride_length = mean(ride_length_minutes),
   number_of_rides = n()
  ) %>% ungroup() %>% arrange(desc(number_of_rides))

#selecting the top 10 rides

top_stations<-head(bike_stations, n= 10)
print(top_stations)
```

# finding the coordinates of the top stations to be represented in tableau
```{r}
#finding their appropriate coordinates to be represented in tableau by merging cyclic_data4 and topstations

popular_stations<-left_join(top_stations,cyclic_data4,by = "start_station_name") %>% arrange(desc(number_of_rides)) %>% group_by(start_station_name) %>% distinct(start_station_name, .keep_all = TRUE)

print(popular_stations)

# saving 
# write_csv(popular_stations,"./popular_stations.csv")
```


# finding the most prefered routes
```{r}

cyclic_data4$routes<- paste(cyclic_data4$start_station_name,"to", cyclic_data4$end_station_name)

length(unique(cyclic_data4$routes))

#showing the top routes
top_routes<-cyclic_data4 %>%group_by(routes) %>% summarise(
  Minutes_total = sum(ride_length_minutes),
  average_ride_length = mean(ride_length_minutes),
  number_of_rides =n()) %>% arrange(desc(number_of_rides)) %>% ungroup()

head(top_routes,n =10)





```


# Act:

## the data shows a higher number of casual riders than active members

## From the data it is observed that the period from june to september shows the most number of active riders with casual members being the most active except in september. since the main aim is to design a marketing campaign to convert casual members to active members. The campaign should focus on this period  

## The marketing team should focus on the 10 top stations presented here. During this period and offer weekly and monthy membership schemes should be offered. 
## Also reducing the membership fees a little might be a good marketing strategy to gain more active members. 

## the marketing campaign could be advertising on popular social media platforms and streamin services like facebook, instagram, twitter, youtube, netflix, amazon prime.

## Posters can be made in this period and posted at strategic venues like near train stations, popular shopping malls etc

## the data also shows that the most popular bike type is classic bike with 64%%, followed by electric bikes with 31% and docked type with 5%

## A recommendation is that further analysis needs to be carried out to investigate what causes the increase in the number of casual riders from June to august.  A possible reason might because it is summer and the weather is sunny. I suggest carrying out a correlation analysis. 


